\begin{thebibliography}{99}

% 1
\bibitem{li-jian-armin-eberlein}
Li~Jian y Armin~Eberlein. 2009. \emph{An analysis of the history of classical software development and agile development}. Proceedings of the 2009 International Conferencen on Systems, Man, and Cybernetics. San Antonio, TX, USA. Octubre 2009. DOI: \url{https://ezproxy.itcr.ac.cr:2878/10.1109/ICSMC.2009.5346888}

% 2
\bibitem{agile-manifesto}
Kent~Beck, Mike~Beelde, Arie~van Bennekum, Allistair Cockburn, et al. 2001. \emph{Manifesto for Agile Software Development}. Agile Alliance.

% 3
\bibitem{version-one}
Version One. 2017. \emph{11th Annual State of Agile Report}. Obtenido de \url{http://stateofagile.versionone.com/}

% 4
\bibitem{henrik-b}
Henrik Bærbak Christensen. 2016. \emph{Teaching DevOps and Cloud Computing using a Cognitive Apprenticeship and Story-Telling Approach}. In Proceedings of the 2016 ACM Conference on Innovation and Technology in Computer Science Education (ITiCSE '16). ACM, New York, NY, USA, 174-179. DOI: \url{https://ezproxy.itcr.ac.cr:2878/10.1145/2899415.2899426}

% 5
\bibitem{hazzan-dubinsky}
Orit Hazzan and Yael Dubinsky. 2007. \emph{Why software engineering programs should teach agile software development}. SIGSOFT Softw. Eng. Notes 32, 2 (March 2007), 1-3. DOI: \url{http://ezproxy.itcr.ac.cr:2075/10.1145/1234741.1234758} 

% 6
\bibitem{hickey-salas}
Timothy J. Hickey and Pito Salas. 2013. \emph{The entrepreneur`s bootcamp: a new model for teaching web/mobile development and software entrepreneurship}. In Proceeding of the 44th ACM technical symposium on Computer science education (SIGCSE '13). ACM, New York, NY, USA, 549-554. DOI:\url{http://ezproxy.itcr.ac.cr:2075/10.1145/2445196.2445361}

% 7
\bibitem{cubric}
Marija Cubric. 2008. \emph{Agile learning \& teaching with wikis: building a pattern}. In Proceedings of the 4th International Symposium on Wikis (WikiSym '08). ACM, New York, NY, USA, , Article 28 , 2 pages. DOI: \url{http://ezproxy.itcr.ac.cr:2075/10.1145/1822258.1822296}

% 8
\bibitem{ding-yousef-yue}
Dabin Ding, Mahmoud Yousef, and Xiaodong Yue. 2017. \emph{A case study for teaching students agile and scrum in Capstone course}. J. Comput. Sci. Coll. 32, 5 (May 2017), 95-101.

% 9
\bibitem{haaranen-lehtinen}
Lassi Haaranen and Teemu Lehtinen. 2015. \emph{Teaching Git on the Side: Version Control System as a Course Platform}. In Proceedings of the 2015 ACM Conference on Innovation and Technology in Computer Science Education (ITiCSE '15). ACM, New York, NY, USA, 87-92. DOI: \url{http://ezproxy.itcr.ac.cr:2075/10.1145/2729094.2742608}

% 10
\bibitem{kropp-meier-2}
Martin~Kropp, Andreas Meier. 2014. \emph{New sustainable teaching approaches in software engineering education}. IEEE Global Engineering Education Conference (EDUCON). IEEE, 2014, pp. 1019–1022. DOI: \url{https://ezproxy.itcr.ac.cr:2878/10.1109/EDUCON.2014.6826229}

% 11
\bibitem{scharlau}
Bruce A. Scharlau. 2013. \emph{Games for teaching software development}. In Proceedings of the 18th ACM conference on Innovation and technology in computer science education (ITiCSE '13). ACM, New York, NY, USA, 303-308. DOI: \url{http://ezproxy.itcr.ac.cr:2075/10.1145/2462476.2462494}

% 12
\bibitem{schroeder-et-al}
Andreas Schroeder, Annabelle Klarl, Philip Mayer, Christian Kroiß. 2012. \emph{Teaching agile software development through lab courses}. Global Engineering Education Conference (EDUCON), IEEE. DOI: \url{https://ezproxy.itcr.ac.cr:2878/10.1109/EDUCON.2012.6201194}

% 13
\bibitem{steghoger-et-al}
Jan-Philipp Steghöfer, Eric Knauss, Emil Alégroth, Imed Hammouda, Håkan Burden, and Morgan Ericsson. 2016. \emph{Teaching Agile: addressing the conflict between project delivery and application of Agile methods}. In Proceedings of the 38th International Conference on Software Engineering Companion (ICSE '16). ACM, New York, NY, USA, 303-312. DOI: \url{https://ezproxy.itcr.ac.cr:2878/10.1145/2889160.2889181}

% 14
\bibitem{advance-it}
Advance IT Minnesota.  2016.\emph{Renewing the IT Curriculum: Responding to Agile, DevOps, and Digital Transformation}. Rep. (November 1, 2016). St. Paul, MN. Obtenido de \url{www.DynamicIT.education}

% 15
\bibitem{cenfotec-2}
Cenfotec. 2016. \emph{``Currículos basados en proyectos integradores.pdf'' y póster ``Proyectos integradores en U Cenfotec.pdf''}. SIIC Simposio Internacional sobre Innovaciones Curriculares, UCR, 2016.

% 16
\bibitem{prosic}
Claudio Pinto, Rafael Herrera, Francisco Mata, Rosaura Matarrita y otros. 2009. \emph{Formación de capital humano en el sector de TIC en Costa Rica}. \hskip 1em plus 0.5em minus 0.4em\relax Facultad Latinoamericana de Ciencias Sociales. Programa de Investigación sobre Economía del Conocimiento en América Latina y el Caribe. Obtenido de \url{http://www.prosic.ucr.ac.cr/sites/default/files/documentos/caphum.pdf}

% 17
\bibitem{trejos-1}
Ignacio Trejos, Alvaro Cordero. 2017. \emph{Learn-by-doing-collaboratively across the curriculum: Integrative projects at UCenfotec}. IEEE World Engineering Education Conference – EDUNINE2017. 3.

% 18
\bibitem{salazar}
Gabriela Salazar. 2012 \emph{Desafíos del curso de ingeniería de software}. Revista de Educación en Ingeniería. Enero a Junio de 2012, Vol 7, Num 13. Páginas 32-43.

% 19
\bibitem{mora-et-al-1}
Sonia Mora, Mayela Coto, Georges Alfaro. 2014. \emph{A proposal for implementing PBL in programming courses}. Computing Conference (CLEI), 2014 XL Latin American. Páginas 1-11. DOI: \url{https://ezproxy.itcr.ac.cr:2878/10.1109/CLEI.2014.6965195}

% 20
\bibitem{mora-et-al-2}
Sonia Mora, Mayela Coto, Georges Alfaro. 2014. \emph{Giving more autonomy to computer engineering students: Are we ready?}. 2013 IEEE Global Engineering Education Conference (EDUCON). Páginas 618-626. DOI: \url{https://ezproxy.itcr.ac.cr:2878/10.1109/EduCon.2013.6530170}


% 21
\bibitem{rashina-et-al}
Rashina~Hoda, Philippe~Kruchten, James~Noble y Stuart~Marshall. 2010. \emph{Agility in context}. SIGPLAN Not. 45, 10 (Octubre 2010), 74-88. DOI: \url{https://ezproxy.itcr.ac.cr:2878/10.1145/1932682.1869467}

% 22
\bibitem{ford}
Neil~Ford. 2011 \emph{Agile Engineering Practices}. Video. O`Reilly Media Inc. Obtenido de \url{https://www.safaribooksonline.com/library/view/neal-ford-on/9781449314439/}. ISBN: 9781449314439

% 23
\bibitem{diebold-dahlem}
Philipp~Diebold y Marc~Dahlem. 2014. \emph{Agile practices in practice: a mapping study}. En Proceedings of the 18th International Conference on Evaluation and Assessment in Software Engineering (EASE `14). ACM, New York, NY, USA, , Artículo 30 , 10 páginas. DOI: \url{http://ezproxy.itcr.ac.cr:2075/10.1145/2601248.2601254}

%24
\bibitem{shore-warden}
James~Shore y Shane Warden. \emph{The Art of Agile Development}. 2008. O'Reilly Media, Inc.

% 25
\bibitem{beck-andres}
Kent~Beck, Cynthia Andres. 2004. \emph{Extreme Programming Explained}. Addison-Wesley Professional.

% 26
\bibitem{bang-et-al}
Soon K. Bang, Sam Chung, Young Choh, y Marc Dupuis. 2013. \emph{A grounded theory analysis of modern web applications: knowledge, skills, and abilities for DevOps}. En Proceedings of the 2nd annual conference on Research in information technology (RIIT `13). ACM, New York, NY, USA, 61-62. \url{DOI=http://ezproxy.itcr.ac.cr:2075/10.1145/2512209.2512229}

% 27
\bibitem{anderson}
Charles~Anderson. 2015. \emph{Docker}. IEEE Software 32, 3 (May-June 2015). 102-c3. 4 páginas.  DOI: \url{https://doi.org/10.1109/MS.2015.62}.

% 28
\bibitem{jabbari-et-al}
Ramtin~Jabbari, Nauman bin Ali, Kai~Petersen y Binish Tanveer. 2016. \emph{What is DevOps?: A Systematic Mapping Study on Definitions and Practices}. En Proceedings of the Scientific Workshop Proceedings of XP2016 (XP `16 Workshops). ACM, New York, NY, USA, Artículo 12 , 11 páginas. DOI: \url{http://ezproxy.itcr.ac.cr:2075/10.1145/2962695.2962707} 

% 29
\bibitem{wiggins}
Adam~Wiggins. \emph{The Twelve-Factor App}. 2017. Obtenido de \url{https://12factor.net}

% 30
\bibitem{puppet-devops}
Puppet Labs, DevOps Research \& Assessment(DORA). 2017. \emph{State of DevOps}. Reporte. Obtenido de \url{https://puppet.com/resources/whitepaper/state-of-devops-report} 


% 31
\bibitem{kropp-meier-1}
Martin~Kropp, Andreas Meier. 2013. \emph{Teaching Agile Software Development at University Level: Values, Management, and Craftsmanship}. En Software Engineering Education and Training (CSEE\&T). Páginas 179 - 188. DOI: \url{https://ezproxy.itcr.ac.cr:2878/10.1109/CSEET.2013.6595249}

% 32
\bibitem{radermacher-walia}
Alex Radermacher and Gursimran Walia. 2013. \emph{Gaps between industry expectations and the abilities of graduates}. In Proceeding of the 44th ACM technical symposium on Computer science education (SIGCSE '13). ACM, New York, NY, USA, 525-530. DOI: \url{http://ezproxy.itcr.ac.cr:2075/10.1145/2445196.2445351}

% 33
\bibitem{murillo-trejos}
Jorge Murillo, Ignacio Trejos. 2017. \emph{IT Skills Mapping Engine}. In World Engineering Education Conference (EDUNINE), IEEE. Vol 2, March 29-22, 2017. Páginas 49-50


% 34
\bibitem{betz-et-al}
Charles Betz, Amos O. Olagunju, and Patrick Paulson. 2016. \emph{The Impacts of Digital Transformation, Agile, and DevOps on Future IT curricula}. In Proceedings of the 17th Annual Conference on Information Technology Education (SIGITE '16). ACM, New York, NY, USA, 106-106. DOI: \url{http://ezproxy.itcr.ac.cr:2075/10.1145/2978192.2978205}

% 35
\bibitem{acm-curriculum}
ACM. 2017. \emph{Curricula Recommendations}. Obtenido de: \url{http://www.acm.org/education/curricula-recommendations}

% 36
\bibitem{acm-cs-curriculum}
Joint Task Force on Computing Curricula, Association for Computing Machinery (ACM) and IEEE Computer Society. 2013. \emph{Computer Science Curricula 2013: Curriculum Guidelines for Undergraduate Degree Programs in Computer Science}. ACM, New York, NY, USA. DOI: \url{http://dx.doi.org/10.1145/2534860}

% 37
\bibitem{acm-se-curriculum}
The Joint Task Force on Computing Curricula. 2015. \emph{Curriculum Guidelines for Undergraduate Degree Programs in Software Engineering}. Technical Report. ACM, New York, NY, USA. 

% 38
\bibitem{acm-msis-curriculum}
Heikki Topi, Helena Karsten, Sue A. Brown, João Alvaro Carvalho, Brian Donnellan, Jun Shen, Bernard C. Y. Tan, and Mark F. Thouin. 2017. \emph{MSIS 2016: Global Competency Model for Graduate Degree Programs in Information Systems}. Technical Report. ACM, New York, NY, USA.

% 39
\bibitem{cenfotec-1}
Cenfotec 2011. \emph{Educación de calidad para la industria de software}. ALTEC 2001, IX Seminario Latino-Iberoamericano de Gestión Tecnológica. Innovación Tecnológica en la Economía del Conocimiento, Costa Rica, 10. 2001.

%40
\bibitem{humble}
Jez Humble, David Farley. 2011. \emph{Continuous delivery: reliable software releases through build, test, and deployment automation}. Pearson Education, Inc.

%41
\bibitem{michael-b}
Michael Hüttermann. 2012. \emph{DevOps for Developers}. Apress.

%%%%%%%% Ordenar aparicion
\bibitem{shackelford-et-al}
Russell Shackelford, Andrew McGettrick, Robert Sloan, Heikki Topi, Gordon Davies, Reza Kamali, James Cross, John Impagliazzo, Richard LeBlanc, and Barry Lunt. 2006. \emph{Computing Curricula 2005: The Overview Report}. In Proceedings of the 37th SIGCSE technical symposium on Computer science education (SIGCSE '06). ACM, New York, NY, USA, 456-457. DOI=http://dx.doi.org/10.1145/1121341.1121482  

\bibitem{rozovsky}
Julia, Rozovsky. \emph{The five keys to a successful Google team. re:Work}. 2015. Obtenido de  \url{https://rework.withgoogle.com/blog/five-keys-to-a-successful-google-team/}

\bibitem{sinaes}
Sistema Nacional de Acreditación de la Educación Superior. \emph{Listado de Carreras Acreditadas}. Obtenido de \url{http://www.sinaes.ac.cr/index.php?option=com_content\&view=article&id=13\&Itemid=115}

\bibitem{creswell}
John Creswell. 2014. \emph{Research Design: Qualitative, Quantitative and Mixed Methods Approaches}. Cuarta Edición. SAGE Publications, Inc.

\end{thebibliography}

